---
title: "GavinKlorfine--Assign4"
author: "Gavin Klorfine"
date: "2026-02-04"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,   # avoid warnings and messages in the output
  message = FALSE,
  fig.height = 5,    # set default figure size
  fig.width = 6.5
)

library(MASS)
library(logmult)
library(ggplot2)
library(dplyr)
library(gklo) # For my gavinTheme() ggplot2 theme. Function included in Appendix
library(vcd)
options(rgl.useNULL = TRUE)
library(vcdExtra)
```

# DDAR Exercise 5.1

### *The data set* `criminal` *in the package* \textsf{logmult} *gives the* $4 \times 5$ *table below of the number of men aged 15–19 charged with a criminal case for whom charges were dropped in Denmark from 1955–1958.*

## Part (a)

### *Use* `loglm()` *to test whether there is an association between* `Year` *and* `Age`*. Is there evidence that dropping of charges in relation to age changed over the years recorded here?*

```{r 5.1a}
data("criminal", package = "logmult")

(mod_ageYr <- loglm(~Age + Year, data = criminal))
```

Due to a significant departure from independence, $\chi^2(`r mod_ageYr$df`) = `r round(mod_ageYr$pearson, 2)`$, $p < .001$, there is evidence that dropping of charges in relation to age changed over the years recorded.

## Part (b)

### *Use* `mosaic()` *with the option* `shade=TRUE` *to display the pattern of signs and magnitudes of the residuals. Compare this with the result of* `mosaic()` *using “Friendly shading,” from the option* `gp=shading_Friendly`*. Describe verbally what you see in each regarding the pattern of association in this table.*

```{r 5.1b}
def_shade <- mosaic(criminal, shade = TRUE, 
                    main ="Criminal Data, Default Shading")
mf_shade <- mosaic(criminal, shade = TRUE, gp = shading_Friendly, 
                   main = "Criminal Data, Friendly Shading")
```

Since most of the residuals lie within the interval $\pm 2$, the default shading scheme does not readily depict a clear pattern regarding cells' relative contributions to the overall $\chi^2$ statistic. In contrast, the mosaic plot with the Friendly shading scheme reveals the sign of cells within the interval $\pm 2$, and consequently, a diagonal pattern of positive residuals. The interpretation here may be that as year (1955-1958) increases, men of younger age (from the interval 15-19) have larger than expected (under independence) frequencies of charges being dropped.


# DDAR Exercise 5.4

### *A related analysis concerns differences in throwing handedness among baseball players according to the fielding position they play. The following code calculates such a frequency table.*

```{r 5.4setup}
library(Lahman)
# NOTE: It seems the "Master" data is now the "People" data
data("People", package = "Lahman")
MasterFielding <- data.frame(merge(People, Fielding, by= "playerID"))
throwPOS <- with(MasterFielding, table(POS, throws))
```

## Part (a)

### *Make a mosaic display of throwing hand vs. fielding position.*
```{r 5.4a}
head(throwPOS)

throwPOS <- throwPOS[,2:3]  # Remove columns B/S
names(dimnames(throwPOS)) <- c("Position", "Throwing Hand")

mosaic(throwPOS, shade = TRUE, 
       main = "Throwing hand Versus Fielding Position\nFrom the People Dataset")
```

## Part (b)

### *Calculate the percentage of players throwing left-handed by position. Make a sensible graph of this data.*

```{r 5.4b}
(pc_left <- 100 * throwPOS[,1] / (throwPOS[,1] + throwPOS[,2]))

pc_left_df <- data.frame(
  Position = names(pc_left),
  throw_left = throwPOS[,1],
  throw_right = throwPOS[,2],
  Percentage_Left = unname(pc_left),
  row.names = NULL
)

pc_left_df <- pc_left_df |> arrange(Percentage_Left)
pc_left_df$Position <- factor(pc_left_df$Position, levels = pc_left_df$Position)

ggplot(pc_left_df, aes(x = Position, y = Percentage_Left)) +
  geom_point() +
  geom_text(
    label = paste0(round(pc_left_df$Percentage_Left, 2), "%"),
    nudge_y = 1.5
  ) +
  coord_cartesian(ylim = c(0, 30)) +
  scale_y_continuous(breaks = seq(0, 30, 5)) +
  labs(
    title = "Percentage of Players Throwing Left-Handed by Position",
    y = "Percentage"
  ) +
  gavinTheme()
  
```

## Part (c)

### *Re-do the mosaic display with the positions sorted by percentage of left-handers.*

```{r 5.4c}
throwPOS_sort <- cbind(   # Create table based on sorted dataframe
  xtabs(throw_left~Position, pc_left_df), 
  xtabs(throw_right~Position, pc_left_df)
)   

colnames(throwPOS_sort) <- c("L", "R")
names(dimnames(throwPOS_sort)) <- c("Position", "Throwing Hand")

mosaic(throwPOS_sort, shade=TRUE, 
       main = "Throwing hand Versus Fielding Position\nFrom the People Dataset")
```

## Part (d)

### *Is there anything you can say about positions that have very few left-handed players?*

Other than 1B, all positions with few left-handed players are in the infield as a consequence of most plays requiring a throw to the infielder's left-hand side (to 1B). The first-baseman does not have this issue as throws would most often be towards their right-hand side (i.e., to 2B and 3B but not home).


# DDAR Exercise 5.8

### *The* `HairEyePlace` *data in* \textsf{vcdExtra} *gives similar data on hair color and eye color, for both Caithness and Aberdeen as a* $4 \times 5 \times 2$ *table.*

## Part (a)

### *Prepare separate mosaic displays, one for each of Caithness and Aberdeen. Comment on any difference in the pattern of residuals.*

```{r 5.8a, fig.height=7, fig.width=7}
data("HairEyePlace", package = "vcdExtra")

mosaic(HairEyePlace[,,"Caithness"], shade = TRUE, gp = shading_Friendly,
       main = "Hair Colour Versus Eye Colour\nin Caithness From the\nHairEyePlace Dataset")

mosaic(HairEyePlace[,,"Aberdeen"], shade = TRUE, gp = shading_Friendly,
       main = "Hair Colour Versus Eye Colour\nin Aberdeen From the\nHairEyePlace Dataset")
```

The pattern of residuals is largely the same, with brightness of eyes/hair being associated with larger than expected frequencies of a similar brightness of eyes/hair (compared to a model of independence). Thus, in both displays there appears to be a pattern of agreement where diagonal elements have large positive residuals, and larger negative residuals tend to be further from the diagonal.

On average, data from Aberdeen exhibit a larger magnitude of residuals than the data from Caithness. The pattern of agreement/disagreement is more evident in the Aberdeen data due to the cells corresponding to medium hair reflecting larger negative residuals (in magnitude).

## Part (b)

### *Construct conditional mosaic plots, using the formula* `~ Hair + Eye | Place` *and both* `mosaic()` *and* `cotabplot()`*. It is probably more useful here to suppress the legend in these plots. Comment on the difference in what is shown in the two displays.*

```{r 5.8b, fig.height=10, fig.width=10}
mosaic(
  ~Hair + Eye | Place, data = HairEyePlace, 
  shade = TRUE, gp = shading_Friendly, legend = FALSE, 
  rot_labels = c(right = -45),
  main = "Model: ~ Hair + Eye | Place"
)
cotabplot(
  ~Hair + Eye | Place, data = HairEyePlace, 
  shade = TRUE, gp = shading_Friendly, legend = FALSE
)
```

The mosaic display allows for a comparison of relative sample size between Caithness and Aberdeen, whereas the cotabplot allows for easier inspection of residuals within the place (Caithness) that has a lower relative sample size.

# Appendix

## gavinTheme()

```{r gavinTheme}
gavinTheme <- function(){
   #' gavinTheme
   #'
   #' Theme for ggplot2 graphs.
   #'
   #' Used by adding to existing ggplot2 object.
   #' @examples
      #' ggplot(...) + gavinTheme()

   theme_minimal() +
   theme(
      plot.title = element_text(hjust = .5, margin = margin(r = 20,b = 10)),
      axis.ticks = element_line(colour = "black", size =.5),
      axis.line = element_line(colour = "black", size = .5),
      axis.text = element_text(size = 10),
      axis.title.x = element_text(margin = margin(t = 10)),
      axis.title.y = element_text(margin = margin(r = 10)),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      plot.margin = margin(t = 10, r = 20, b = 10, l = 10)
   )
}
```