---
title: "GavinKlorfine--Assign2"
author: "Gavin Klorfine"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,   # avoid warnings and messages in the output
  message = FALSE,
  fig.height = 5,    # set default figure size
  fig.width = 6.5
)
options(rgl.useNULL = TRUE)
library(vcdExtra)
library(dplyr)
library(xtable)
library(ggplot2)

#' My R package with my ggplot theme function; I have included the function 
#' below so you are able to run this on your system
library(gklo)
```

```{r function, include=FALSE}
gavinTheme <- function(){
   #' gavinTheme
   #'
   #' Theme for ggplot2 graphs.
   #'
   #' Used by adding to existing ggplot2 object.
   #' @examples
      #' ggplot(...) + gavinTheme()

   theme_minimal() +
   theme(
      plot.title = element_text(hjust = .5, margin = margin(b = 10)),
      axis.ticks = element_line(colour = "black", size =.5),
      axis.line = element_line(colour = "black", size = .5),
      axis.text = element_text(size = 10),
      axis.title.x = element_text(margin = margin(t = 10)),
      axis.title.y = element_text(margin = margin(r = 10)),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank()
   )
}
```

# DDAR Exercise 2.3

*The data set `UCBAdmissions` is a 3-way table of frequencies classified by `Admit`, `Gender`, and `Dept`.*

**(a)** *Find the total number of cases contained in this table.*
```{r 2.3a}
data("UCBAdmissions")

sum(UCBAdmissions)
```

**(b)** *For each department, find the total number of applicants*
```{r 2.3b}
ucb <- as.data.frame(UCBAdmissions)

totals <- (
  ucb |> 
  group_by(Dept) |> 
  summarize(s = sum(Freq))
)

totals
```

**(c)** *For each department, find the overall proportion of applicants who were admitted*
```{r 2.3c}
admsns <- (
  ucb |>
  group_by(Dept, Admit) |>
  filter(Admit == "Admitted") |>
  summarize(admit_freq = sum(Freq))
)

admsns$total_freq <- totals$s
admsns$ovr_prop <- admsns$admit_freq / admsns$total_freq
admsns |> select(Dept, ovr_prop)
```

**(d)** *Construct a tabular display of department (rows) and gender (columns) showing the proportion of applicants in each cell who were admitted relative to the total applicants in that cell*
```{r 2.3d, results='asis'}
berk_tab <- (
  prop.table(UCBAdmissions, c("Dept", "Gender"))
)

berk_tab <- berk_tab["Admitted",,] |> t() |> xtable()
caption(berk_tab) <- "Proportion of admitted applicants by department and gender using the UCBAdmissions dataset."
berk_tab
```

# DDAR Exercise 3.1

*The `Arbuthnot` data in HistData (Example 3.1) also contains the variable `Ratio`, giving the ratio of male to female births.*

**(a)** *Make a plot of `Ratio` over `Year`, similar to Figure 3.1. What features stand out? Which plot do you prefer to display the tendency for more male births?*
```{r 3.1a}
# Code adapted from Friendly & Meyer (2016; p. 67), Discrete Data Analysis with R

data("Arbuthnot", package= "HistData")

with(Arbuthnot, {
   plot(x = Year, y = Ratio, type= "b",
   ylim= c(1.0, 1.17), main = "Arbuthnot's Data on Sex Ratios in London",
   ylab= "Ratio of Males:Females")
   abline(h= 1.0, col= "red", lwd= 2)
   abline(h= mean(Ratio), col= "blue")
   lines(loess.smooth(Year, Ratio), col= "blue", lwd= 2)
   text(x = 1660, y= 1.0, expression(H[0]: "# Male Christenings = # Female Christenings"),
   pos= 3, col= "red")
})
```

Features that stand out include no points dipping below 1.0, the value expected if there were an equal number of males and females christened. The average ratio of males:females christened from `r min(Arbuthnot$Year)` to `r max(Arbuthnot$Year)` was `r round(mean(Arbuthnot$Ratio), 3)`; this average ratio tended to be larger from `r min(Arbuthnot$Year)` to ~1670 (ratio = `r round(mean(Arbuthnot$Ratio[Arbuthnot$Year < 1670]), 3)`), whereas it tended to be smaller from ~1670 to `r max(Arbuthnot$Year)` (ratio = `r round(mean(Arbuthnot$Ratio[Arbuthnot$Year > 1670]), 3)`). Despite these fluctuations, the ratio remained near the mean, indicating a consistent trend of more male births than female births.

I prefer the plot in Figure 3.1 as it better represents the underlying process (i.e., births as Bernoulli trials where there is a probability that the child will be male). Probabilities rather than ratios also allow for modelling via a binomial distribution.

**(b)** *Plot the total number of christenings, Males + Females or Total (in 000s) over time. What unusual features do you see?*
```{r 3.1b}
# Code adapted from Friendly & Meyer (2016; p. 67), Discrete Data Analysis with R

data("Arbuthnot", package= "HistData")

with(Arbuthnot, {
   plot(x = Year, y = Total, type= "b",
   ylim= c(0, 17), main = "Arbuthnot's Data on Total Number of Christenings in London",
   ylab= "Total Number of Christenings (in Thousands)")
   abline(h= mean(Total), col= "blue")
   lines(loess.smooth(Year, Total), col= "blue", lwd= 2)
})
```

An unusual feature is that of the data between ~1645 and 1660, as they form a trough in what would otherwise be a steady increase in total christenings. Furthermore, the descent into the trough is somewhat gradual, occurring over ~10 years, whereas the ascent from the trough is quick and occurs over ~5 years.

There is also a data point near ~1705 that seems atypical, as the number of christenings is a few thousand less than the surrounding data points (i.e., 12,000 as compared to a typical value of 15,000 for the previous 20 years).


# DDAR Exercise 3.3

*Use the data set `WomenQueue` to:*

**(a)** *Produce plots analogous to those shown in Section 3.1 (some sort of bar graph of frequencies).*

```{r 3.3a, fig.height = 4}
data("WomenQueue")
wq <- as.data.frame(WomenQueue)

ggplot(wq, aes(x = nWomen, y = Freq)) +
  geom_bar(
    stat ="sum",
    fill = "steelblue",
    color = "black"
  ) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_x_discrete(expand = c(0, 0)) +
  labs(
    title = "Number of Women in 100 10-Person Queues",
    x = "Number of Women in Queue",
    y = "Frequency"
  ) +
  gavinTheme()
```

**(b)** *Check for goodness-of-fit to the binomial distribution using the `goodfit()` methods described in Section 3.3.2.*

```{r 3.3b_stand}
wq_fit <- goodfit(WomenQueue, type = "binomial")
summary(wq_fit)

plot(wq_fit, type="standing", 
     main = "Women Queue Data, Fitting a Binomial Model",
     xlab = "Number of Women in Queue"
)
```

```{r 3.3b_hang}
plot(wq_fit, type="hanging", shade = TRUE,
     main = "Women Queue Data, Fitting a Binomial Model",
     xlab = "Number of Women in Queue"
)
```

```{r 3.3b_dev}
plot(wq_fit, type="deviation", shade = TRUE,
     main = "Women Queue Data, Fitting a Binomial Model",
     xlab = "Number of Women in Queue"
)
```

Due to a non-significant *p* value and no residuals exceeding $\pm \: 2$, $\text{Bin}(10, `r unlist(wq_fit$par)[1]`)$ can be considered a good fit for the data.

**(c)** *Make a reasonable plot showing departure from the binomial distribution.*

```{r 3.3c, fig.height = 6}
distplot(
  WomenQueue, 
  type = "binomial",
  xlab = "Number of Women in Queue",
  main = "Binomialness plot: WomenQueue data"
)
```

**(d)** *Suggest some reasons why the number of women in queues of length 10 might depart from a binomial distribution,* $\text{Bin}(n=10, p=1/2)$

Because the true probability of selecting a woman from a queue of 10 may differ from $p = 1/2$ (in the place and time data was collected). Perhaps there were more men taking the subway or there were more men in London, generally.

# DDAR Exercise 3.4
*Continue Example 3.13 on the distribution of male children in families in Saxony by fitting a binomial distribution,* $\text{Bin}(n = 12, p = \frac{1}{2})$*, specifying equal probability for boys and girls. [Hint: you need to specify both `size` and `prob` values for `goodfit()`.]*

**(a)** *Carry out the GOF test for this fixed binomial distribution. What is the ratio of $\chi^2 / df$? What do you conclude?*

```{r 3.4a}
data("Saxony")

# GOF test
sax_fit <- goodfit(
  Saxony, 
  type = "binomial",
  par = list(size = 12, prob = 0.5)
)
sum_sax <- summary(sax_fit)

# Ratio of chisq/df
(chi_df <- sum_sax["Pearson", "X^2"] / sum_sax["Pearson", "df"])
```

I conclude that $\text{Bin}(n = 12, p = \frac{1}{2})$ is a significantly bad fit to the `Saxony` data, $\chi^2(`r sum_sax["Pearson", "df"]`) = `r round(sum_sax["Pearson", "X^2"], 3)`$, $p < .001$. Furthermore, the ratio of $\chi^2 / df = `r round(chi_df, 3)`$ is substantially greater than 1, which would be the expected value under the null hypothesis of acceptable fit.

**(b)** *Test the additional lack of fit for the model* $\text{Bin}(n = 12, p = \frac{1}{2})$ *compared to the model* $\text{Bin}(n = 12, p = \hat{p})$ *where* $\hat{p}$ *is estimated from the data.*

```{r 3.4b}
sax_fit_phat <- goodfit(
  Saxony, 
  type = "binomial",
  par = list(size = 12)
)
sum_sax_phat <- summary(sax_fit_phat)

phat <- unlist(sax_fit_phat$par)["prob"]
```

While the likelihood ratio improved (i.e., lowered) by estimating $\hat{p}$ from the data ($G^2_{\hat{p} = `r round(phat,3)`}(`r sum_sax_phat["Likelihood Ratio", "df"]`) = `r round(sum_sax_phat["Likelihood Ratio","X^2"], 2)`$ versus $G^2_{p = 0.5}(`r sum_sax["Likelihood Ratio", "df"]`) = `r round(sum_sax["Likelihood Ratio","X^2"], 2)`$), there was still a significant departure from the null hypothesis of good fit (*p* < .001), meaning that $\text{Bin}(n = 12, \hat{p} = `r round(phat, 3)`)$ is still a very bad fit for the `Saxony` data.

**(c)** *Use the `plot.goodfit()` method to visualize these two models.*

```{r 3.4c}
plot(
  sax_fit, 
  type = "deviation",
  shade = TRUE,
  main = "Saxony Data, Fitting Bin(n = 12, p = 1/2)",
  xlab = "Number of Males"
)   # 3.4a

plot(
  sax_fit_phat,
  type = "deviation",
  shade = TRUE,
  main = paste0("Saxony Data, Fitting Bin(n = 12, p = ", round(phat, 2), ")"),
  xlab = "Number of Males"
)   # 3.4b
```


