---
title: "GavinKlorfine--Project1_Supplemental"
author: "Gavin Klorfine"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,   # avoid warnings and messages in the output
  message = FALSE,
  fig.height = 6,    # set default figure size
  fig.width = 6.5
)

as_table <- function(obj, freq = NULL, dims = NULL){
  
  # If user supplied a table or array, return it back to them
  if (length(intersect(c("table", "array"), class(obj))) > 0){
    
    tab <- as.table(obj) # Handle arrays
    
    # To include dimensions if specified
    if (!is.null(dims)){
      tab <- margin.table(tab, margin = dims)
    }
    
    return(tab)
  }
  
  
  # If obj is a tibble, convert to data frame
  else if (length(intersect("tbl", class(obj))) > 0){
    obj <- as.data.frame(obj)
  }
  
  if (!is.null(dims)){ # If dims supplied by user, use those
    cols <- dims
  }
  else { # If dims NOT supplied by user, use everything else
    cols <- colnames(obj)
  }
  
  if (!is.null(freq)){  # If freq supplied by user, then... (freq form)
    cols <- cols[cols != freq] # Remove freq column
    tab <- xtabs(reformulate(cols, response = freq), data = obj) # freq ~ cols
  }
  else { # If freq NOT supplied by user, and not array, then... (case form)
    tab <- xtabs(reformulate(cols), data = obj)
    
    # Check if user forgot to supply freq, warn if they potentially forgot
    common <- c("n", "freq", "frequency", "count")
    if (length(intersect(tolower(colnames(obj)), common)) > 0){
      warning("Ensure a value for 'freq' was supplied if your data was in frequency form.")
    }
  }
  return(tab)
}

options(rgl.useNULL = TRUE)
library(vcdExtra)
library(MASS)
library(ca)
library(ggplot2)
library(dplyr)
library(car)
library(effects)
```

# Question 1

### **Minnesota High School Graduates:** The `Hoyt` data in \textsf{vcdExtra} gives a $4 \times 3 \times 7 \times 2$ table classifying nearly 14000 graduates by (a) post-high school `Status`, (b) `Rank` in their graduating class, (c) father's  `Occupational` status (7 levels, from 1=High to 7=Low) and (d) `Sex`.  `Status` is considered the outcome variable.  How does it depend on the other factors?

## Part (a)

### *Analyze and display the associations between status and each of the other variables individually. In doing this, you might also consider focusing attention on the distinction between College and Non-college Status.*

```{r q1a_occupation_analyze}
data("Hoyt", package = "vcdExtra")

(status_occup <- (margin.table(Hoyt, c("Status", "Occupation")) |> CMHtest())) # row means

# Analyze college vs non-college
distinct_status <- collapse.table(
  Hoyt, Status = c("College", "Non-college", "Non-college", "Non-college")
)

(
  dist_status_occup <- (
    margin.table(distinct_status, c("Status", "Occupation")) |> CMHtest() # row means
  )
)
```

```{r q1a_occupation_display, fig.show="hold", out.width="50%"}
mosaic(
  ~ Occupation + Status, data = Hoyt,
  shading = TRUE, gp = shading_Friendly,
  rot_labels = c(top = 45),
  main = "Graduate's Post High School Status\nand Father's Occupational Status"
)
mosaic(
  ~ Occupation + Status, data = distinct_status,
  shading = TRUE, gp = shading_Friendly,
  main = "With Post High School Status Collapsed\nto College/Non-College"
)
```

```{r q1a_rank_analyze}
(status_rank <- margin.table(Hoyt, c("Status", "Rank")) |> CMHtest()) # row means

(
  dist_status_rank <- (
    margin.table(distinct_status, c("Status", "Rank")) |> CMHtest() # row means
  )
)
```

```{r q1a_rank_display, fig.show="hold", out.width="50%"}
mosaic(
  ~ Rank + Status, data = Hoyt,
  shading = TRUE, gp = shading_Friendly,
  rot_labels = c(top = 45),
  main = "Graduate's Post High School Status\nand Rank in Their Graduating Class"
)
mosaic(
  ~ Rank + Status, data = distinct_status,
  shading = TRUE, gp = shading_Friendly,
  main = "With Post High School Status Collapsed\nto College/Non-College"
)
```

```{r q1a_sex_analyze}
(status_sex <- margin.table(Hoyt, c("Status", "Sex")) |> assocstats())

(
  dist_status_sex <- (
    margin.table(distinct_status, c("Status", "Sex")) |> assocstats()
  )
)
```

```{r q1a_sex_display, fig.show="hold", out.width="50%"}
mosaic(
  ~ Sex + Status, data = Hoyt,
  shading = TRUE, gp = shading_Friendly,
  rot_labels = c(top = 45),
  main = "Graduate's Post High School Status\nand Their Sex"
)
mosaic(
  ~ Sex + Status, data = distinct_status,
  shading = TRUE, gp = shading_Friendly,
  main = "With Post High School Status Collapsed\nto College/Non-College"
)
```

## Part (b)

### *Consider and analyze the associations among the predictor variables: *`Rank`*, *`Occupation` *and* `Sex`*. Is there anything of interest worth mentioning or analyzing in further detail? Is it useful to consider collapsing some of the categories?*

```{r q1b_beforecollapse, fig.width=6.5}
# Display associations
pred_tab <- margin.table(Hoyt, c("Rank", "Occupation", "Sex"))

pairs(pred_tab, gp = shading_Friendly)
```

```{r q1b_aftercollapse}
# Collapsing occupation to the following based on similarity in the mosaic plots:
# 1, 2 -> "High",   |         4 -> "Low"
#    3 -> "Middle"  |   5, 6, 7 -> "Very Low"

coll.pred_tab <- collapse.table(
  pred_tab,
  Occupation = c("High", "High", "Middle", "Low", "Very Low", "Very Low", "Very Low")
)

# Associations among predictors
CMHtest(margin.table(coll.pred_tab, c("Rank", "Occupation"))) # Test nonzero correlation
CMHtest(margin.table(coll.pred_tab, c("Rank", "Sex"))) # Test difference in column means
CMHtest(margin.table(coll.pred_tab, c("Sex", "Occupation"))) # Test difference in row means

pairs(coll.pred_tab, gp = shading_Friendly)
```


## Part (c)

### *Start by fitting a simple model of joint independence of* `Status` *from the predictors, then add additional associations as you consider necessary until you obtain a reasonably well-fitting model.*

```{r q1c, fig.height = 10}
coll_Hoyt <- collapse.table(
  Hoyt, 
  Occupation = c("High", "High", "Middle", "Low", "Very Low", "Very Low", "Very Low")
)

(joint <- loglm(~Rank*Occupation*Sex + Status, data = coll_Hoyt))

(rankstat <- loglm(~Rank*Occupation*Sex + Rank*Status, data = coll_Hoyt))

(occupstat <- loglm(~Rank*Occupation*Sex + Status*(Rank + Occupation), data = coll_Hoyt))

(sexstat <- loglm(~Rank*Occupation*Sex + Status*(Rank + Occupation + Sex), data = coll_Hoyt))

(statsexocc <- loglm(~Rank*Occupation*Sex + Status*(Rank + Occupation + Sex) + Status*Sex*Occupation, data = coll_Hoyt))

(statsexrank <- loglm(~Rank*Occupation*Sex + Status*(Rank + Occupation + Sex) + Status*(Sex*Occupation + Occupation*Rank), data = coll_Hoyt))


anova(joint, rankstat, occupstat, sexstat, statsexocc, statsexrank)


mosaic(statsexrank, rot_labels = c(right = -45))
```

## Part (g)

### *Conduct a multiple correspondence analysis for the* `Hoyt` *data. Provide a reasonable interpretation of the two dimensions that explain the largest proportion of variance.*

```{r q1g, fig.width=7, fig.height=7}
mjca.coll_Hoyt <- mjca(coll_Hoyt, lambda = "Burt")
summary(mjca.coll_Hoyt)

### BELOW ADAPTED FROM Friendly & Meyer (2016; p. 245), Discrete Data Analysis with R ###
res <- plot(mjca.coll_Hoyt, labels= 0, pch= ".", cex.lab= 1.2)

# extract factor names and levels
coords <- data.frame(res$cols, mjca.coll_Hoyt$factors)
nlev <- mjca.coll_Hoyt$levels.n
fact <- unique(as.character(coords$factor))

cols <- c("blue", "red", "darkgreen", "black")
points(coords[,1:2], pch=rep(16:19, nlev), col=rep(cols, nlev), cex=1.2)
text(coords[,1:2], label=coords$level, col=rep(cols, nlev), pos=3, cex=1.2, xpd=TRUE)

legend(
  "bottomright", legend= c("Status", "Rank", "Occupation", "Sex"), title= "Factor", 
  title.col= "black", col = cols, text.col = cols, pch= 16:19, bg= "gray95", cex = 1.2
)
```

# Question 2

### **Student Opinion about the Vietnam War:** *The data set *`Vietnam` *in* \textsf{vcdExtra} *gives a* $2 \times 5 \times 4$ *contingency table in frequency form reflecting a survey of student opinion on the Vietnam War at the University of North Carolina in May 1967. The table variables are *`sex`*, *`year` *in school and *`response`*, which has categories: (A) Defeat North Vietnam by widespread bombing and land invasion; (B) Maintain the present policy; (C) De-escalate military activity, stop bombing and begin negotiations; (D) Withdraw military forces Immediately. How does the chosen response vary with *`sex` *and* `year`*?*

## Part (a)

### *Construct some useful plots showing the proportions of each response category by* `sex` *and* `year` *of study. It may be helpful to add some smoothed or fitted curves to aid interpretation.*

```{r q2a}
data("Vietnam", package = "vcdExtra")

viet <- rename(Vietnam, "Sex" = sex, "Year" = year, "Response" = response)

# Male proportions sum to 1 and female proportions sum to 1 within each response category
respProps_viet <- (
  viet |> 
  group_by(Response, Sex) |> 
  mutate(Proportion = Freq / sum(Freq))
)

ggplot(respProps_viet, aes(x = Year, y = Proportion, fill = Sex, colour = Sex)) + 
  facet_wrap(vars(Response)) + 
  geom_point() + 
  geom_line() +
  scale_y_continuous(breaks = seq(0, 0.6, 0.1)) +
  coord_cartesian(ylim = c(0, 0.6)) +
  labs(
    title = "Proportions of Each Response Category by Sex and Year",
    subtitle = "Response"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    panel.background = element_blank(),
    panel.border = element_rect()
  )

viet_tab <- as_table(viet, freq = "Freq") # See Appendix for as_table()
cotabplot(
  ~ Sex + Year | Response, data = viet_tab, 
  shade = TRUE, gp = shading_Friendly,
)
```

## Part (b)

### *Analyze the associations between* `response` *and each of* `sex` *and* `year` *separately, using graphs and statistical tests. Note that both* `year` *and* `response` *can be considered ordinal factors. *

```{r q2b}
# Create tables
tab.viet_sex <- margin.table(viet_tab, margin = c("Sex", "Response"))
tab.viet_year <- margin.table(viet_tab, margin = c("Year", "Response"))

# Statistical tests for association
CMHtest(tab.viet_sex) # Test row means
CMHtest(tab.viet_year) # Test nonzero correlation

# Graphically analyze associations
mosaic(tab.viet_sex, shade = TRUE, gp = shading_Friendly)
mosaic(tab.viet_year, shade = TRUE, gp = shading_Friendly)
```

## Part (c)

### *Fit loglinear and/or polytomous logit models designed explain the relationship between response and year of study*

```{r q2c}
viet$Response <- ordered(viet$Response) # Order by "level of pacifism": A < B < C < D

polr.viet_intrcpt <- polr(Response ~ 1, data = viet, Hess = TRUE)
polr.viet <- polr(Response ~ Year, data = viet, Hess = TRUE)

summary(polr.viet)

anova(polr.viet_intrcpt, polr.viet) # Year not significant predictor over and above intercept only model

plot(Effect("Year", polr.viet), main = "Effect Plot for the Effect of\nYear of Study on Response")



loglm.viet_simple <- loglm(~(Year*Sex) + Response, data = viet_tab) # Predictors independent of response

loglm.viet_respYear <- loglm(~(Year*Sex) + Response*Year, data = viet_tab) # 2-way of response*year

loglm.viet <- loglm(~(Year*Sex) + Response*(Year + Sex), data = viet_tab) # All 2-ways--good fit

loglmComp.viet <- anova(loglm.viet_simple, loglm.viet_respYear, loglm.viet) # Model comparison


mosaic(loglm.viet, gp = shading_Friendly)
```

## Part (g)

### *Conduct a multiple correspondence analysis for the* `Vietnam` *data. Provide a reasonable interpretation of the two dimensions that explain the largest proportion of variance.*

```{r q2g}
mjca.viet_tab <- mjca(viet_tab, lambda = "Burt")
summary(mjca.viet_tab)

### BELOW ADAPTED FROM Friendly & Meyer (2016; p. 245), Discrete Data Analysis with R ###
res <- plot(mjca.viet_tab, labels= 0, pch= ".", cex.lab= 1.2)

# extract factor names and levels
coords <- data.frame(res$cols, mjca.viet_tab$factors)
nlev <- mjca.viet_tab$levels.n
fact <- unique(as.character(coords$factor))

cols <- c("blue", "red", "black")
points(coords[,1:2], pch=rep(16:18, nlev), col=rep(cols, nlev), cex=1.2)
text(coords[,1:2], label=coords$level, col=rep(cols, nlev), pos=3, cex=1.2, xpd=TRUE)

legend(
  "bottomright", legend= c("Sex", "Year", "Response"), title= "Factor", 
  title.col= "black", col = cols, text.col = cols, pch= 16:18, bg= "gray95", cex = 1.2
)
```

# Appendix

## `as_table()`

Converts object (`obj`) in frequency or case form into table form. The column containing the frequencies (`freq`) must be supplied if `obj` is in frequency form.

```{r as_table}
as_table <- function(obj, freq = NULL, dims = NULL){
  
  # If user supplied a table or array, return it back to them
  if (length(intersect(c("table", "array"), class(obj))) > 0){
    
    tab <- as.table(obj) # Handle arrays
    
    # To include dimensions if specified
    if (!is.null(dims)){
      tab <- margin.table(tab, margin = dims)
    }
    
    return(tab)
  }
  
  
  # If obj is a tibble, convert to data frame
  else if (length(intersect("tbl", class(obj))) > 0){
    obj <- as.data.frame(obj)
  }
  
  if (!is.null(dims)){ # If dims supplied by user, use those
    cols <- dims
  }
  else { # If dims NOT supplied by user, use everything else
    cols <- colnames(obj)
  }
  
  if (!is.null(freq)){  # If freq supplied by user, then... (freq form)
    cols <- cols[cols != freq] # Remove freq column
    tab <- xtabs(reformulate(cols, response = freq), data = obj) # freq ~ cols
  }
  else { # If freq NOT supplied by user, and not array, then... (case form)
    tab <- xtabs(reformulate(cols), data = obj)
    
    # Check if user forgot to supply freq, warn if they potentially forgot
    common <- c("n", "freq", "frequency", "count")
    if (length(intersect(tolower(colnames(obj)), common)) > 0){
      warning("Ensure a value for 'freq' was supplied if your data was in frequency form.")
    }
  }
  return(tab)
}
```

